Predictive Models to Support Quoting of Fixed Fee Consulting Projects

### 2016-12-16 finalwriteup/Consulting Profitability Paper/Consulting Profitability Paper.Rmd
finished method in paper

Neural Netowrks and SVM were not purusued because black box
A previously mentioned disadvantage is that the NN algorithm is a black box because the internal structure is too complex for interpretation. They also require a lot of training data relative to other methods.

Also, SVM's can be very slow to train and therefore not suitable for industry purposes [@Auria2008]. Because it is so important to engage decision makers with a model that can explain its results, these two methods are not appropriate for the effort estimation problem.

@@

binary classification problems, where the objective is to predict the probability of an event occurring. Probability is a continuous response variable, however predictions from a linear regression would not be bound by 0 and 1.

@@

Because of the linear relationship, an equation can be fit to the log odds of the binary response variable against each of the explanatory variables using linear regression. Then, the linear equation describing the log odds can be transformed back to probability by taking the inverse log i.e. the exponential and rearranging [@Macdonald1975]:

The result in a sigmoidal function bound by 0 and 1. The sigmoidal function originates from a linear fit of the log odds in the data. Assume linear relationship between covariates and response variable

 A coefficient represents the change in the log odds of the response variable for each unit increase in an explanatory variable. Therefore, taking the exponential of the coefficient is the change in odds of the response variable. If the change in odds was 2, then if the explanatory variable increased by 1, the response variable event would be twice as likely to occur.

 Logistic Regression is a good  benchmark to compare other binary predictive models due to its simplicity and speed [@Moore1989]. single linear relationship means Not prone to overfitting

@@

single decision trees also suffer from low predictive accuracy and instability, so to combat these problems, ensemble tree methods were pioneered in the 1990's [@Breiman1996].

The three methods of combining hundreds or thousands of trees turn decision trees into high performing, stable predictors.

training multiple trees from the same data by sampling bootstrapped training sets with replacement. However, when creating each tree, a random subset of attributes (variables) is considered at each split. The reduced subset of attributes is resampled for each split in the tree. This allows dominant variables to be suppressed for a fraction of the splits, allowing the algorithm to explore signals in weaker variables [@Breiman2001a].

Additionally, ensemble tree methods are faster than SVM's and Neural Networks but can perform just as well and even provide insights such as variable importance and variable relationships [@Sealfon2012].

@@

Lastly, the boosted decision tree approach applies gradient descent theory to a series of decision trees. The trees are limited to a certain depth to maintain simplicity, and each tree models the residuals (or errors) of the preceding tree.
By modeling the errors, misclassified cases are weighted higher than correctly classified cases and influence the structure of the latest tree more [@Hastie2009]. This increased weighting on errors is why the algorithm is called boosting.
The limited depth of each tree prevents overfitting at each stage and the combined result of up to thousands of trees is very powerful [@Elith2008].

@Caruana2006 tested Boosted Trees, Random Forests, Neural Networks, SVM's, Logistic Regression and Naive Bayes on 11 binary classification problems and found that Boosted Trees performed best, followed by Random Forests. This demonstrates how ensemble tree methods are capable of competing with high-level machine learning algorithms.

Trees are not built on a probabilistic framework, and therefore their results cannot be provided in this framework [@Louppe2014].

@@

The Naive Bayes method works by making conditional independence assumptions about the explanatory variables in order to simplify probability calculations for the response variable (the response variable must be categorical).

These equations are simple to compute given the data, and the class with the highest probability can then be chosen [@Provost2013].

The advantages of this method are that the conditional independence assumption enables very fast calculations and predictions. The method can perform well in real world tasks because the assumption of independence does not significantly damage predictions [@Provost2013]. Assumptions of independence genearlly not true (correlated variables). This is fine for ranking but the output probabilities are not accurate statistical probabilities [@Caruana2006].

Naive Bayes method generally provides a good benchmark to compare against more complex models that should outperform it [@Caruana2006].

##### 3.1.5 Bayesian Networks

A Bayesian Network is a graphical probabilistic model that illustrates the conditional dependencies between variables in a data set. The model is visually represented by a directed acyclic graph (DAG) and is capable of linking the conditional dependency between any variable to another variable [@Heckerman1998]. The conditional relationships are Bayesian, where the probabilities in one node are conditional upon values in nodes directed towards it as well as preceding nodes

Bayesian Networks reduce computations required to find the probability of a unique combination given other explanatory variable values while still taking into account many conditional dependencies that Naive Bayes [@Barber2012]

Naive Bayes is the simplest form of a Bayesian Network [@Zhang2004]

Network relationships can be learned from the data, however this is not widely included as part of the suite of machine learning methods. The conditional variable dependencies are calculated from the data, which in turn can define the graph structure. Some conditional dependencies can be set before the structure is learned [@Barber2012].

Drawbacks of Bayesian Networks include the difficulty for them to process continuous variables.

Bayesian networks have found success in combining deterministic models with observed data as well as expert knowledge.[@Kragt2009]

@@

 It includes first how the data was obtained, the lengthy cleaning process, 
 Projects varied from total invoiced amounts of \$500 to over \$1,000,000
 a project could have thousands of rows of relevant data that needed to be converted into a single row per project.
 10 engineered variables describing timesheets: Percent of hours performed by 'professional' employees as opposed to 'technical' employees, Position of the employee that completed the most hours on each project
 3 engineered from invoicing data: 
 4 from project data: number of projects completed with each client
 text analysis of project titles to develop project classifications

 Discretisation was performed by generating a hierarchical dendrogram of each variable to visualise the clusters.

Limiting the predictive model to a concise set of meaningful variables reduces noise and improves predictions. Less variables and a simpler model is easier for stakeholders to understand [@Weisberg2005]. followed by variable importance analysis, variable selection: varialbe importance aalyses done with linear regression, random forest, cforest to represent the range of algroithms that were trialled. 11 variables chosen:

 trials of selected predictive algorithms.  binary classification problem predicting profit or loss.  Complex models should be measured against simple models that can be built at a fraction of the computational cost.

 * Regression - baseline model
* Naive Bayes - baseline model
* Bayesian Network - grabage.
* Random Forest
* Gradient Boosted Trees

To compare the models, the area under the receiver operating characteristic (ROC) curve (AUC) statistic was used. An ROC graph visualises the curve from which an AUC score is calculated. In this problem a 'positive' is a loss making job. For binary classification, AUC is a more meaningful statistic than classification accuracy when the output is a probability that can be applied to the problem. The AUC indicates model performance across many thresholds while classification accuracy represents a single threshold.

a model must be first created using a training set. The model can then make predictions on the test set. this is used to build the ROC
A model that is perfectly classified would have an AUC = 1.
A model that predicts as well as random chance would have an AUC = 0.5. An AUC between 0.5 and 1 means the model is performing better than random chance.

In order to compare which models performed significantly better than others, an adequate sample size of results statistics is required. Multiple models could be made by using different data in the training vs. testing sets, each providing a resulting test statistic (RMSE or AUC).  Initially, 20 models of each method were created in this fashion. Then a two-sample power calculation was run using the two sets of 20 results to determine the sample size to achieve a statistical power of 0.8 (100 to achieve power of 0.8 between boosted trees and naive bayes)

Missing Data Imputation. All methods except gradient Boosted Trees and Naive Bayes could not handle missing data. a complete data set allows for complete sets of predictions from each method, and these predictions could then be blended to further improve results. The MICE Random Forest method was chosen for imputation because it has been proven to work well with complex data sets [@Shah2014]. If similar predictive results were obtained using Boosted Trees imputed data and imputed data, the imputed data must be reasonable. The imputed data set was then trialed on the remaining methods and compared to unimputed trials.

 Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Six blending methods, ranging from simple to complex, were tested using predictions from the top performing individual models. These included simple averaging of the individual model results, building a Logistic Regression model using the individual model results only, a Boosted Tree model using individual model results only, feature weighted linear stacking (FWLS), Random Forests, and Boosted Trees. A simplified explanation of FWLS is as a Linear Regression where meta-features as well as model predictions from individual models are included as explanatory variables. Then, each meta-feature is interacted with each set of model results [@Sill2009].
 Feature interaction is performed passively due to the nature of how trees are built. A split in a node determined by one variable is conditional upon the preceding split, which was based on another variable and so on.
 The simplest method averaged the probability output (a number between 0 and 1) of all three models. The next two simplest methods consisted of building a Logistic Regression and Boosted Trees model from the three prediction model outputs only. Feature Weighted Linear Stacking (FWLS), Random Forest and Boosted Trees were also tested as blending methods in a more complex scenario where the predictions from each model became additional variables to the original explanatory variables (called meta-features in this context). The model predictions were interacted with each original variable so that models that performed more strongly under certain meta-feature states could be weighted as such.

Again, 100 models were built to achieve a power of 0.8 for the variation in AUC test statistics across different divisions of training/testing data. A maximum of five blended models were created from a complete set of test results (built from the individual models), then a new set of test results were created for the next five blended models.

 Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision-making scenarios. This analysis differed from accuracy in predicting profitability, so several blended models as well as individual models were carried forward for this analysis.  This was done using a profit curve - a chart that plots the change in profit the company earns on the y-axis vs. the probability threshold on the x-axis. A simple approach was taken for this analysis, where projects with a probability to be a loss making job greater than the threshold were rejected entirely. Therefore all profits and losses from jobs above the threshold were discounted. An equation defining the change in profit as a percentage of the original profit using threshold rejection is shown below:

\begin{equation}
\Delta\ Overall\ Profit\ \textrm{(\%)} = \frac{\sum_{p = 1}^{N} I(Pr_p < threshold) \cdot Profit_p}{\sum_{p = 1}^{N} Profit_p} \cdot100
\end{equation}

@@

If the threshold was zero, all jobs were rejected and the profit would be $0. If the threshold was 1.0, all jobs were accepted and the profit would be the same as the profit the company actually experienced since the data is a sample of historic projects. The aim was to find the optimal threshold point where saying 'no' to a job above that level would result in higher profits, because jobs that were likely to make a loss were being rejected. This chart will clarify what percentage of profit increase the company could expect by integrating the algorithm into decision-making.

the curve varies with different divisions of the data. Therefore, in order to understand the uncertainty around the profit curve and to determine which blended or individual model performed statistically better than others, a large sample size of curves was required. 100 to achieve a power of 0.8. which was achieved by repeating 5-fold training/testing splits. A 95% confidence interval could also be determined around the highest point on each curve. The final expected increase in profit and the percentage of projects to be rejected presented a clear scenario that the case study business managers could assess in terms of their business strategy.

>>>>

A range of simple and complex, linear and non-linear methods were applied to the problem including Logistic Regression, Random Forests, Gradient Boosted Trees, and Naiive Bayes. Furthermore, logistic regression, random forests, boosted trees and simple averaging were used to blend predictions from individual models. A brief description of each method is provided below.

<!-- Neural Netowrks and SVM were not purusued because black box -->
<!-- A previously mentioned disadvantage is that the NN algorithm is a black box because the internal structure is too complex for interpretation. They also require a lot of training data relative to other methods. -->

<!-- Also, SVM's can be very slow to train and therefore not suitable for industry purposes [@Auria2008]. Because it is so important to engage decision makers with a model that can explain its results, these two methods are not appropriate for the effort estimation problem.  -->

@@

The aim of this research is to build a model that predicts whether a consulting project will be profitable or not. Therefore each method must be capable of binary classifcation. Binary classification output from a model is the probability, between 0 and 1, of an event occuring. In this case, the event is a project being loss-making.

@@

Logisitic regression models a linear relationship bewteen the log odds of a binary response variable and each of the explanatory variables. This is performed similarly to linear regression, but is transformed to probability of an event occuring by taking the inverse log of the log odds and rearranging [@Macdonald1975]:

$$
\begin{aligned}
\log(odds) &= \beta_0 + X_1\beta_1 \\
\log\left(\frac{p}{1-p}\right) &= \beta_0 + X_1\beta_1 \\
-\log\left(\frac{1-p}{p}\right) &= \beta_0 + X_1\beta_1 \\
-\log\left(\frac{1}{p} - 1\right) &= \beta_0 + X_1\beta_1 \\
\log\left(\frac{1}{p} - 1\right) &= -\beta_0 - X_1\beta_1 \\
\exp\left(\log\left(\frac{1}{p} - 1\right)\right) &= \exp(-\beta_0 - X_1\beta_1) \\
\frac{1}{p} - 1 &= \exp(-\beta_0 - X_1\beta_1) \\
\frac{1}{p} &= 1 + \exp(-\beta_0 - X_1\beta_1) \\
p &= \frac{1}{1 + \exp(-\beta_0 - X_1\beta_1)}
\end{aligned}
$$

The result in a sigmoidal function bound by 0 and 1. A coefficient represents the change in the log odds of the response variable for each unit increase in an explanatory variable. Logistic Regression is a good  benchmark to compare other binary predictive models due to its simplicity and speed [@Moore1989]. Addittionally, because of the simple, single linear relationship, the method is not prone to overfitting [@Perlich2003].

@@

Ensemble tree methods were pioneered in the 1990's to combat the low predictive accuracy and instability of individual decision trees. Combining hundreds or thousands of trees transform decision trees into high performing, stable predictors [@Breiman1996].

The Random Forests algorithm works by training multiple trees from the bootstrapped samples of a dataset. Additionally, when creating each tree, a random subset of attributes (covariates) is considered at each split. The reduced subset of covariates is resampled for each split in the tree. Dominant covariates are then suppressed for a fraction of the splits, allowing the algorithm to explore signals in weaker covariates [@Breiman2001a]. Performance statistics and tests run within the algorithm provide insights to the modeller, such as variable importance and variable relationships [@Sealfon2012].

@@

The boosted decision tree approach applies gradient descent theory to a series of decision trees. Performance tends to be higher if individual trees are limited to a certain depth to maintain simplicity and prevent overfitting [@Elith2008]. Unlike Random Forests, the trees are sequential, where each tree models the residuals (or errors) of the preceding tree.

By modeling the errors, misclassified cases are weighted higher than correctly classified cases and influence the structure of the next tree more [@Hastie2009]. This increased weighting of error cases is called boosting. The successive results of up to thousands of trees is very powerful [@Elith2008]. @Caruana2006 tested Boosted Trees, Random Forests, Neural Networks, SVM's, Logistic Regression and Naive Bayes on 11 binary classification problems and found that Boosted Trees performed best, followed by Random Forests. This demonstrates how capable ensemble tree methods are of competing with high-level machine learning algorithms.

@@

Naive Bayes is a simple, fast method included in this work as it provides a good baseline method against which to compare complex methods [@Caruana2006]. It works by making conditional independence assumptions about the explanatory variables in order to simplify probability calculations for a categorical response variable. For example, if predicting a binary variable, $R$ (with classes 0 or 1) given two covariates $X_1$ and $X_2$, the equations for the probability of both outcomes is as follows:

$p(R = 1|X_1 X_2) = p(X_1|R = 1) \cdot p(X_2|R = 1) \cdot p(R = 1)$

$p(R = 0|X_1 X_2) = p(X_1|R = 0) \cdot p(X_2|R = 0) \cdot p(R = 0)$

These equations are simple to compute given the data, and the class with the highest probability can then be chosen [@Provost2013]. The method can perform well in real world tasks because the assumption of independence does not significantly damage predictions, but amplifies the magnitude of the probability [@Provost2013]. This is fine for ranking class 0 against 1 but the output probabilities are not accurate statistical probabilities [@Caruana2006].

##### 3.1.5 Model Blending

Several research groups involved in the high profile data science competition, Netflix Prize, developed sophisticated methods of model stacking, otherwise known as ensemble methods or model blending [@Sill2009]. In this paper, the term 'blending' will be used to avoid confusion with ensemble tree methods. Model blending combines predictions from several models that used methods with different theoretical foundations. In this way, the strengths of each method can be combined. Since the 1990's research in statistics has shown that averaging results from different methods can provide better predictive accuracy than any single model [@Madigan1994].

Six blending methods, ranging from simple to complex, were chosen. These included simple averaging of the individual model results, building a Logistic Regression model and Boosted Tree model using individual model results only, feature weighted linear stacking (FWLS), Random Forests, and Boosted Trees. The first two models are simply the average, and weighted average of predictions from separate individual models. For example, if the Logistic Regression model output a probability of 0.6 for one case, while a Boosted Tree output 0.7, a simple blended average for that case would be a probability of 0.65. The last three methods are more complex, and take advantage of potentially meaningful interactions between the individual model predictions and the original covariates describing each case (called meta-features in the model blending literature) [@Sill2009]. For example, if the Logistic Regression model predicted profitability better than the Boosted Tree model for projects with only one team member, the complex blending methods would take advantage of this interaction. In summary, in the complex methods, the probability outputs from each model are interacted with each of the original covariates. Model blending has the potential to elevate output from the original methods by combining each of their strengths.

@@

The raw data from the case study company's database consisted of three separate datasets: one describing the client and technical details for each project, another with each employee's time sheet entries, and another with all past invoices. This data was combined, cleaned, and rearranged so that each project was described within one row. From the timesheet data, 10 variables were engineered including the percent of hours performed by 'professional' employees as opposed to 'technical' employees, and the position of the employee that completed the most hours on each project. 3 variables were engineered from the invoicing data such as total amount invoiced for a project and mean invoice size for a client. Finally, from the project dataset, 2 variables were engineered including the number of projects completed with a client, and project category. Text analysis of the project description determined the project category variable. The response variable, whether the project was profitable or not, was calculated by subtracting the cost of employee hours from the timesheets from the sum of the paid invoices for a project. The total amount invoiced for projects ranged from \$500 to over \$1,000,000.

34 covariates made up the full dataset, and these were narrowed down to 11 in order to reduce predictive noise and provide a simple model that is comprehendable for stakeholders [@Weisberg2005]. Variable selection methods included linear regression, random forests, and cforests and the final variables are listed below.

* % of Hours Completed by a Professional-level Employee
* Time span
* Number of Employees on Project
* Amount Invoiced for Project
* Business Category of Client
* Project Category
* Total Amount Invoiced from the Client - Past Jobs
* Discipline
* Position of Main Project Employee - new variable from ANOVA findings
* % of Hours by Main Project Employee
* Billing Type

Timespan and the total amount invoiced are unknowns at before a project commences. Therefore, these variables were discretised into 5 or 6 categories from which a manager would be able to select. The category boundaries were determined by visualising clusters using hierarchical dendrograms.

##### 3.2.1 Research Question 1

To answer the first research question, if loss making projects can be predicted using statistical and machine learning algorithms, the binary classification and blending methods were built and evaluated. The area under the receiver operating characteristic (ROC) curve (AUC) statistic was used to compare models and assess performance. Models were built on a training subset of data (80%) and validated on the remaining data. Therefore, different training and testing sets produced different AUC statistics. In order to compare which models performed significantly better than others, an adequate sample size of AUC statistics was required. A two-sample power calculation run on 20 AUC output for each model determined that 100 samples were required to achieve a statistical power of 0.8.

Then, the model output from the validation sets were combined and fed into the three simple and three complex blending models. Again, 100 models were built to achieve a power of 0.8 for the variation in AUC test statistics across different divisions of training/testing data. A maximum of five training and validation divisions were created from a complete set of individual model results.

##### 3.2.2 Research Question 2

The second research questions asks whether a predictive algorithm can improve a company's profitability? This was analysed using a simple decision-making scenario. The model with the highest AUC score may not improve company profits the most, because this depends on *which* loss making and profitable projects are accurately being identified. Therefore, all blended models as well as individual models were carried forward for this analysis. A profit curve was built to compare methods, which is a chart that plots the change in profit the company earns on the y-axis vs. the probability threshold on the x-axis. For this decision making scenario, the threshold is the probability output value (a value between 0 and 1 that the model predicts for each case), from which it is decided whether a project will be accepted or rejected. For example, if the output threshold is 0.6, and a model predicts a project has probability of being a loss-making job of 0.7, the project is not taken on by the case study company. All profits and losses from jobs above the threshold were discounted. The y-axis of the profit curve is the ratio of profit after rejecting projects based on model output to historical profit:

$$
\begin{aligned}
Ratio\ of\ Model\ Profit\ to\ Historical\ Profit\ (\%) &= \frac{\sum_{p = 1}^{N} I(Pr_p < threshold) \cdot Profit_p}{\sum_{p = 1}^{N} Profit_p} \cdot100
\end{aligned}
$$

@@

If the threshold were zero, all jobs were rejected, the profit would be $0, and the ratio to historical profits would be 0%. If the threshold was 1.0, all jobs were accepted and the profit would be the same as historical profits, giving a ratio of 100%. The aim was to find the optimal threshold point where mostly loss-making projects were rejected, resulting in higher profits. 100 profit curves were built for each method to achieve a power of 0.8 and a 95% confidence interval could also be determined around each threshold point. The final expected increase in profit combined with the percentage of projects to be rejected presents a clear scenario that the case study business managers could strategically assess for implementation.
